---
title: "Workflow diary"
date: 2023-04-22
author: Your Name
format: 
  html:
    toc: true
    toc-depth: 3
    code-tools: true
citation: true
---

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

## How to use this template 

The idea of a workflow diary is inspired by lab diaries that are commonly used, for example, in chemistry labs to document goals, setup and outcomes of experiments. 
Similar to a lab diary, every time that you work on your modelling project, record what your goal was, what you tried and what the results were. 

The work sessions can be of varying length and intensity. 
There is no need to stick to the amount of work sessions in this template. 
Find your own rhythm of progressing in your project and record your work sessions accordingly. 

The main goal of this seminar is for you to engage with the modelling workflow and the workflow diary is a way for you to see what you tried and document your questions that can then be discussed during the in-person sessions. 

## Introduction to Bayesian Workflows

### Work session 1 

I set up Quarto and installed the required packages. 
I chose the dataset ... from ... and I want to answer the following questions: 

1. ...
2. ...

The outcome that I am interested in is ... . 

```{r}
data <- ChickWeight
head(data)
```

### Work Session 2 

I visualised the data.

## Choosing an Initial Model

### Work Session 3 

The first model that I tried 

```{r}
```

### Work Session 4

```{r}
```

## Prior Choices

### Work Session 5 

At first, I chose the priors ... because ... . 
Then, I tried ... . 

```{r}
```

### Work Session 6 

### Work Session 7 

## Model Checking with Posterior Predictive Checks & Calibration

### Work Session 8 

I visualised the posterior results for the model parameters and observed ... .

### Work Session 9 
I compared the observed data with replicates from the posterior and observed ... . 

```{r}
```

### Work Session 10 

I read about visualisations for posterior predictive checking in ... . 

## Extending Models and Model Selection

### Work Session 11 

I explored different extension of my initial model(s) by adding ... and changing ... . 

### Work Session 12 

I compared the predictive performance of the different models and concluded that ... . 

```{r}
```

## Interpreting and Presenting Model Results

### Work Session 13 

```{r}
```

### Work Session 14 

At the seminar, I asked about ... and tried out ... . 

### Work Session 15 

## Summary & Presentations

### Work Session 16 

```{r}
```

### Work Session 17 

### Work Session 18 