---
title: "Workflow diary"
date: 2023-04-22
author: Your Name
format: 
  html:
    toc: true
    toc-depth: 3
    code-tools: true
citation: true
---

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

### Session 1: Introduction to Bayesian Workflows

I chose the dataset ... from ... and I want to answer the following questions: 

1. ...
2. ...

The outcome that I am interested in is ... . 

```{r}
data <- ChickWeight
head(data)
```

### Session 2: Choosing an Initial Model

The first model that I tried 

```{r}
```

### Session 3: Prior Choices

At first, I chose the priors ... because ... . 
Then, I tried ... . 

```{r}
```

### Session 4: Model Checking with Posterior Predictive Checks & Calibration

I visualised the posterior results for the model parameters and observed ... .
I compared the observed data with replicates from the posterior and observed ... . 

```{r}
```

### Session 5: Extending Models and Model Selection

I explored different extension of my initial model(s) by adding ... and changing ... . 
I compared the predictive performance of the different models and concluded that ... . 

```{r}
```

### Session 6: Interpreting and Presenting Model Results

```{r}
```

### Session 7: Summary & Presentations

```{r}
```
